% Redefine section to suppress numbering and make it bigger and bold
\titleformat{\section}[block]
  {\normalfont\LARGE\bfseries}{\thesection}{1em}{}
\titlespacing*{\section}{0pt}{*3}{*4}


\chapter{Supplementary Information}
\newpage


\clearpage
\section{Chapter 3.1 -- Integration of diverse bioactivity data into the Chemical Checker compound universe}
\input{sections/Appendix/Protocols}



\clearpage
\section{Chapter 3.2 -- Clustering, visualizing and exploring the chemical space of small molecules using the Chemical Checker signatures}
\input{sections/Appendix/Navigation}



\clearpage
\section{Chapter 3.3 -- Stereochemically-aware bioactivity descriptors for uncharacterized chemical compounds}
\input{sections/Appendix/Stereoisomers}

\clearpage
\section{Chapter 3.4 -- Comprehensive detection and characterization of human druggable pockets through novel binding site descriptors}
\input{sections/Appendix/PocketVec}



\chapter{Publications}
\newpage
\section{List of publications}
\label{ListOfPublications}

1. Fernández-Torras, A., \textbf{Comajuncosa-Creus, A.}, Duran-Frigola, M. and Aloy, P., 2022. Connecting chemistry and biology through molecular descriptors. Current Opinion in Chemical Biology, 66, p.102090.

2. \textbf{Comajuncosa-Creus, A.}, Lenes, A., Sánchez-Palomino, M., Dalton, D. and Aloy, P., 2024. Stereochemically-aware bioactivity descriptors for uncharacterized chemical compounds. Journal of Cheminformatics, 16(1), p.70.

3. \textbf{Comajuncosa-Creus, A.}, Jorba, G., Barril, X. and Aloy, P., 2024. Comprehensive detection and characterization of human druggable pockets through binding site descriptors. Nature Communications, 15(1), pp.1-20.

4. Scholes, N.S., Bertoni, M., \textbf{Comajuncosa-Creus, A.}, Kladnik, K., Frommelt, F., Hinterndorfer, M., Razumkov, H., Prokofeva, P., Schwalm, M.P., Imrichova, H., Barone, E., Schätz, C., Rukavina, A., Koren, A., Kubicek, S., Knapp, S., Gray, N.S., Superti-Furga, G., Kuster, B., Aloy, P., and Winter, G.E., 2024. Inhibitor-induced supercharging of kinase turnover via endogenous proteolytic circuits. bioRxiv. https://doi.org/10.1101/2024.07.10.602881v2.

\newpage
\section{Attachment of publications}

Only the relevant publications for this work are attached below, which correspond to publications #1, #2 and #3 (see \hyperref[ListOfPublications]{List of publications}). 

