\chapter{Contributions}
\label{Contributions}
\clearpage

As stated in the Objectives section, this thesis aims to explore and further extend the concept of small molecule and protein druggable binding site descriptors, thus laying at the intersection between Cheminformatics and Bioinformatics. Accordingly and, outlined per each chapter, we have made the following contributions:

\begin{enumerate}

\item \textbf{Chapter 3.1}: The implementation of the (preexisting) CC data integration protocol, which represents an optimal framework for generating new bioactivity spaces based on user-provided data, to build novel transcriptomics, proteomics, antibiotics and MoA-based bioactivity signatures. 

\item \textbf{Chapter 3.2}: The use of chemical and bioactivity CC signatures to cluster, visualize and navigate the chemical space of small molecules. In addition, the illustration of the potential of bioactivity signatures to perform such tasks in a biologically relevant manner.  

\item \textbf{Chapter 3.3}: The quantification of bioactivity differences, in terms of protein-ligand binding, between different stereoisomers and the development of stereochemically-aware bioactivity signaturizers. 

\item \textbf{Chapter 3.4}: The development of a novel strategy, named PocketVec, to represent and characterize protein druggable binding sites in the shape of numerical vectors (descriptors). PocketVec descriptors have been comprehensively benchmarked, showing excellent performances in multiple tasks while overcoming the main limitations of existing methods. Additionally, PocketVec descriptors have been generated for all pockets included within human protein domains (using experimentally-determined and accurately-predicted protein structures) and have been proven useful to provide complementary insights to classical structure- and sequence-based comparisons as well as to identify experimentally observed protein similarities in terms of protein-compound binding. 

\end{enumerate}