
%%% CONTRIBUTORS %%%

\textbf{Adrià Fernández-Torras} participated in the development of Fig \ref{Introduction_Fig2} and Fig \ref{Introduction_Fig3} (\hyperref[introduction]{Introduction}).

\textbf{Martino Bertoni}, \textbf{Miquel Duran-Frigola}, \textbf{Oriol Guitart-Pla}, \textbf{Martina Locatelli}, \textbf{Yasmmin Martins} and \textbf{Nicolas Soler} designed and developed the CC data integration pipeline (\hyperref[Chapter_3.1]{Chapter 3.1} -- Integration of diverse bioactivity data into the Chemical Checker compound universe).

\textbf{Martino Bertoni}, \textbf{Adrià Fernández-Torras}, \textbf{Nils Kurzawa}, \textbf{Elena Pareja-Lorente}, \textbf{Gema Rojas-Granado} and \textbf{Eva Viesi} participated in the building of the different novel CC bioactivity spaces (\hyperref[Chapter_3.1]{Chapter 3.1} -- Integration of diverse bioactivity data into the Chemical Checker compound universe, particularly B1.002, D1.002 and M1.001).

\textbf{Aksel Lenes} and \textbf{Miguel Sánchez-Palomino} designed and trained the collection of deep neural networks consistuting the Signaturizers3D (\hyperref[Chapter_3.3]{Chapter 3.3} -- Stereochemically-aware bioactivity descriptors for uncharacterized chemical compounds).

Most computational calculations performed in this work were executed on the in-house \textbf{SBNB} cluster, the \textbf{IRB} cluster, or the \textbf{BSC} infrastructure (particularly in StarLife), due to their advanced HPC capabilities.

\textbf{Large Language models}, including but not limited to \textbf{ChatGPT}, have been used for various purposes such as text rephrasing (as applied in this manuscript), English translation and coding. In line with standard recommendations in the field\cite{porsdam_mann_guidelines_2024}, we present the following statement:

\textit{"Any use of generative AI in this manuscript adheres to ethical guidelines for use and acknowledgement of generative AI in academic research. Each author has made a substantial contribution to the work, which has been thoroughly vetted for accuracy, and assumes responsibility for the integrity of their contributions."}

The organization and presentation of this work have been greatly inspired by the doctoral theses of \textbf{Adrià Fernández-Torras} and \textbf{Miquel Duran-Frigola}, former PhD students of the SBNB lab at the IRB. Additionally, this manuscript was developed based on the PhD Thesis and Dissertation \textbf{LaTeX Templates} available on \textbf{Overleaf}, specifically those designed for Harvard University, Princeton University, and New York University (NYU).

\hl{The cover of this thesis was designed by} \textbf{Bryan Saldivar-Espinoza}. 